|

### 2021-12-13 06-Chp-IPD-Analysis.Rmd
Range of edits incl: - Better figure captions - Better description of proportional biases - Julian comments to Discussion

Individual participant data (IPD) meta-analyses are a special type of meta-analysis where those undertaking the study contact the authors of relevant studies to request access to the raw data on each participant. This is different to the meta-analyses presented in the previous chapter, which used published summary results.

@@

To examine the relationship between lipid levels and dementia risk, I applied for access to `r n_applied$total` unique data sources containing information on these variables. However, only a small proportion of these data sources (`r fmt_applied(n_accessed$total)`) provided the requested access.

@@

The reasons for the low response rate to requests for data are explored. Finally, I suggest future studies to investigate how data access rates could be improved.

@@

Individual participant data (IPD) meta-analyses are considered to be the gold standard form of evidence synthesis, allowing for the application of a common selection model and analytical approach across all identified cohorts.[@riley2010] They are particularly useful when investigating the impact of participant-level characteristics, something that is not possible with aggregate data unless the results are stratified by the characteristic of interest.[@riley2010;@thompson2005] Knowledge of which groups a treatment will benefit most (or harm least) is a core aim of the move towards personalised medicine.[@riley2020;@hingorani2013]
IPD analysis also offer a mechanism by which previously unanalysed datasets can be incorporated into an analysis, thus expanding the evidence base for a particular research question.

@@

As such, the aims of this analysis are two fold: firstly, to perform an IPD meta-analysis across identified cohorts to examine the impact of participant age-at-measurement and sex on the effect of lipids on dementia risk; and secondly, to expand the evidence base for the effect of lipids on dementia outcomes by obtaining estimates from previously unanalysed cohorts available via the Dementia Platform UK, a large consortia of dementia cohorts.

<!-- TODO Read this[@riley2020] -->

@@

<!-- TODO Should this be reported in relation to the PRISMA-IPD statement -->

@@

Eligible data sources for this analysis were prospective cohort studies. Data sources which were cross-sectional, either by design or due to the available data (e.g. a study conducted across multiple waves, but only data from a single wave could be accessed) were excluded. Similarly, studies making use of population-level electronic health records, which often require an extensive project proposals in order to gain access to the data, were ineligible due to the time and cost involved in applying. No restrictions were put on the number of participants or the length of follow-up, though participants must be free of dementia (or assumed to be free, based on age at entry)  at baseline.

@@

I considered four blood lipid fractions as part of this analysis, namely: total cholesterol (TC), low-density lipoprotein cholesterol (LDL-c), high-density lipoprotein cholesterol (HDL-c) and triglycerides (TG). Cohorts were eligible for inclusion if they contained data on at least one of these in a continuous format (i.e. studies with binary "hypercholesteromemia" would be excluded).

@@

### Applying for data access

Potentially eligible data sources as defined above were identified via two mechanisms, each with a distinct focus. For both approaches, the number of cohorts responding to the request for data access, and where applicable, the reasons given for a refusal were recorded.

@@

The second approach focused on incorporating relevant, previously-unanalysed data into the analysis, thus providing additional evidence on the relationship between blood lipids and dementia risk. This was achieved through the Dementia Platform UK (DPUK), a collaborative grouping of existing dementia cohorts established by the MRC which works with data owners to make their data readily accessible for secondary analysis.[@bauermeister2020] It provides access to 42 cohorts with over 3 million participants, and makes use of a central streamlined application processs for all cohorts with the intent of making it easier to access data from existing data sources.

@@

Where data on one of these exposures (TC, LDL-c, HDL-c, TG) was missing, it was inferred from the other three fractions (where available) using the Friedwald formula (Equation \@ref(eq:total-cholesterol-formula)). Lipids levels were retained as continuous variables rather than dichotomising into a binary hypercholesterolemia exposure, given the additional statistical power this adds to the meta-analysis.[@riley2020;@ensor2018] Where lipid measurements were reported in _mg/dL_, these were converted to _mmol/L_.

@@

Missing data in this analysis was classified as either relating to missing values (a cohort collected data on a variable of interest, but some values are missing) or missing variables (a cohort did not collect data on a variable of interest). Variables with missing values were identified, and 20 imputed datasets were created.[@sterne2009] Imputation was performed using the MICE (Multiple Imputation by Chained Equations) using the `mice` package in R.

Missing variables were originally intended to be addressed using a previously described method.[@fibrinogenstudiescollaboration2009] Here, the correlation between the fully-adjusted and partially-adjusted estimate in cohorts with the full set of covariates is used to estimate the fully-adjusted effect in those cohorts missing covariates. However, this method requires several large cohorts to contain the full set of covariates, a condition this analysis failed to meet given the very low response rate. As such, two other common approaches were employed.[@fibrinogenstudiescollaboration2009] In the primary analysis, all cohorts were analysed adjusted for the set of common covariates across cohorts (Model 1: age, sex, smoking, alcohol, diabetes). As a sensitivity analysis, cohorts with the full complement of covariates (Whitehall II and EPIC) were analysed using a maximally-adjusted model (Model 2: Model 1, further adjusted for education, ethnicity, and prevalent ischemic heart disease and BMI). Results between the common-set-adjusted (Model 1) and maximally-adjusted (Model 2) models were then compared.

@@

In this analysis, a two-stage IPD analysis was planned. Under a two-stage approach, estimates for the effect of each lipid fraction on incident dementia were first calculated for each data source. More specifically, a logistic regression model adjusted for relevant covariates as detailed in the above section was used to quantify the effect of a 1 _mmol/L_ increase in each lipid fraction on each dementia outcome. Results were expressed as odds ratios (OR). Examination of the data available via the DPUK indicated that the common time-to-event approach used in studies of dementia outcomes would be precluded by the absence of detailed time-to-event data. An overall effect estimate was then produced by combining data-source-specific estimates in a random-effects meta-analysis (see Section \@ref(meta-analysis-methods) for a broader discussion of meta-analysis methods).

@@

A two-stage IPD approach was employed for a number of reasons. Firstly, and most importantly, a two-stage approach allows for siloed data, enabling researchers who are unwilling/unable to provide their data to analyse themselves following a specified analysis plan.[@riley2010] Secondly, it allows for the production of forest plots of within-cohort estimates, something which is useful for the triangulation exercise in Chapter \@ref(tri-heading). Finally, a two-stage approach is simpler to model, as it uses standard well-documented summary effect estimate meta-analysis techniques, and automatically accounts for methodological issues such as clustering within cohorts[@abo-zaid2013] and the potential for ecological bias.[@burke2017]

<!-- TODO Need to state that using a fixed effect model for the interaction terms -->

@@

In order to investigate the interaction of participant-level characteristics (age and sex) with lipid levels, interaction terms for lipid-covariate terms were extracted and synthesised using a fixed effects meta-analysis.[@fisher2017] To avoid ecological bias, where the between-study association does not reflect the within-study associations, cohorts where there was no within-study variation in the covariate of interest were excluded from the interaction analysis for that covariate.[@burke2017] As an example, it is impossible to estimate the impact of sex on the lipid~dementia relationship in a study which contains only female participants. All interaction analysis were performed using the common-set-adjusted model (Model 1).

<!-- QUESTION   -->

@@

<!-- Need exact number from each approach, plus detailed exclusion reasons -->

Of the `r n_applied$total` studies to which I applied for data access, only three (`r paste0(comma((n_accessed$total/n_applied$total)*100),"%")`) were included in the final analysis. Figure \@ref(fig:cohortFlowchart) details whether the cohorts eventually included in the review were identified by the systematic review or via DPUK portal. In addition, the reasons for cohorts not being included in the analysis are presented, where available, stratified by application approach.

@@

(ref:cohortFlowchart-cap) Flowchart of included cohorts, stratified by identification method (systematic review vs DPUK).

(ref:cohortFlowchart-scap) Flowchart

@@

The European Prospective Investigation of Cancer (EPIC) - Norfolk is a population-based cohort, containing men and women recruited from 35 general practices in Norfolk between 1993 and 1998.[@riboli1997; @riboli2002] Data provided by this cohort for the analysis, dementia was ascertained at at the 5th Health Checkup (2016-2018), giving between 18-25 years of follow-up. The added evidential value of the EPIC cohort is small, given the the fact that the data obtained contains only 8 dementia events. The cohort contains no information on dementia subtype, while all covariates of interest bar _Apo_$\mathcal{E}4$ were available.

@@

The Whitehall II study is a prospective cohort study of men and women recruited between 1985 and 1989 from the civil service in London.[@marmot2005] Lipids measurements were available from the third wave of the study, conducted between 1991-1993. The cohort is linked with the Hospital Episode Statistics (HES) database, a database containing details of participant events at NHS hospitals in England, which was used to define covariates at baseline and dementia outcomes.[@zotero-15403] HES data for dementia outcomes was avaialle up to March 2015, providing between 22-24 years of follow-up. Information was available on dementia type, which was classified as all-cause dementia, Alzheimer's disease or vascular dementia. This data source contained details on all covariates of interest except for _Apo_$\mathcal{E}4$ status.

Of note, the Whitehall II cohort was analysed in one of the included studies identified by the systematic review presented in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading),[@tynkkynen2018] meaning that a comparison between the published result and the analysis reported here was was possible.

@@

A key consideration in the definition of covariates across cohorts was the classification of age. The Whitehall II study, the largest cohort to which I had access, only shared age data in five-year age band (e.g. 40-44, 45-49, etc.). To ensure comparability across the cohorts, I created identical categories in the CaPS and EPIC data. This grouped age variable was then used in all subsequent analyses.

@@

<!-- TODO Note here about how the IHD variable in CAPS has a note saying there might be an issue with the variables, while the Whitehall study uses narrow range of codes to define CHD. EPIC is selfreported angina/mi/arrythmia -->

@@

The results from the main effect analysis across the varying lipid fractions on each dementia,  considered can be seen in Figures \@ref(fig:mainEffectDem) & \@ref(fig:mainEffectVad), respectively. There was weak evidence for an association of any lipid level with either all-cause or vascular dementia, with the exception of a harmful association between raised triglycerides and vascular dementia (`r ipd_vasdem`).

<!-- REVIEW Present main effect for WhiteHall AD analysis - or rather highlight that it is shown in Figure XXX in comparison with Tynken -->

@@

Given minimal effect of further adjustment for education and BMI on the effect estimates, the results from the interaction analysis presented here were obtained using the common-set-adjusted model. Further, while the maximally-adjusted model (Model 2) could have been employed in the analysis of the effect of sex, as the CaPS cohort contains only men and had to be excluded from this analysis, employing the same underlying model helped to ensure comparability between the age and sex interaction estimates.

@@

<!----------------------------------------------------------------------------->
(ref:interactionDementiaAge-cap)  Meta-analysis of interaction terms representing the effect of a one-step increase in age group on the lipid/all-cause dementia association

(ref:interactionDementiaAge-scap) shortcap

@@

<!----------------------------------------------------------------------------->
(ref:interactionDementiaSex-cap) Meta-analysis of interaction terms representing the effect of male gender on the lipid/all-cause dementia association

(ref:interactionDementiaSex-scap) shortcap

@@

<!----------------------------------------------------------------------------->
(ref:interactionVascularAge-cap) Meta-analysis of interaction terms representing the effect of a one-step increase in age group on the lipid/vascular dementia association

(ref:interactionVascularAge-scap) shortcap

@@

This analysis requested data from `r n_applied$total` data sources, but only obtained data from three, all of which were based in the United Kingdom. No evidence for an effect of lipids on the risk of dementia or related outcomes was identified, except for a harmful association of raised triglycerides with risk of vascular dementia. Similarly, there was weak evidence for an interaction of the effect of lipids levels on dementia outcomes with participants' age, grouped into 5 year bands, or sex.

@@

The obvious key limitation of this analysis is the very low response rate to requests for data access, which may bias the results if there are systematic differences in the association of lipids with dementia outcomes between cohorts that share data and those that do not.[@ahmed2012] Whether or not to press ahead with an IPD analysis in the absence of all (or even most) data is a personal decision, and some previous analyses have highlighted where they decided not to pursue an IPD analysis.[@jaspers2014] For the purposes of this thesis, the decision was made to conduct the IPD analysis, as it provided training in application of the methods in addition to providing additional evidence sources that could be incorporated into the triangulation exercise detailed in Chapter \@ref(tri-heading).

@@

A low response rate is not unexpected, given that a review of IPD studies published between 1987 and 2015 found that fewer than half managed to obtained data from greater than 80% of studies, and that in many cases, the exact percentage of studies for which data was obtained was not accurately reported.[@nevitt2017] However, it is assumed that a ~10% response rate is at the lower end of the scale.

There are likely several likely reasons for this low response rate to requests for data access. In general terms, there are a several well-documented barriers that prevent data from being made readily available, including concerns regarding participant privacy, fear of "scooping" or "parasitic" behaviour, and a lack of trust between primary and secondary researchers.[@vanpanhuis2014] More specific to this analysis, IPD meta-analysis including studies other than randomised controlled trials have less success in obtaining IPD from studies.[@nevitt2017] Additionally, while no evidence is available on whether the characteristics of the researcher requesting data access influences the response rate and eventual decision, there is the possibility that my position as a PhD student meant I was less likely to elicit a positive response than a well-known senior academic might. The timing of the requests for data access, coinciding with a global pandemic, may also have affected the response rate as researchers prioritised COVID-related work.

Finally, the method of contact used (email) has been shown to be less successful in eliciting responses from authors when compared with telephoning.[@danko2019] One of the reasons for this may be that the email addresses reported on publications are more likely to be out of date for older publications. Anecdotally, post-hoc investigation of a subset of cohorts revealed that several corresponding/first authors were no longer at the same institution as when the study was reported, and as a result, were unlikely to have access to the institutional email address listed on the study publication.

The obstacles to data access described above are in theory what the DPUK was built to address. However, even with the help of the streamlined application process afforded by the DPUK, accessing sufficient data was a challenge. The response rate among DPUK cohorts a year after application was just 50%. In addition, some cohorts responded that that the proposed study question was already under investigation by another group, and that they would not share the data on this basis. In light of this, a centralised database of ongoing analysis being performed in DPUK would be of enormous help. Finally, the DPUK process would be aided by a clearer distinction between those cohorts that are "DPUK native" (i.e. where a copy of the data is already held on DPUK servers) versus externally hosted, as the response time for externally hosted cohorts is likely to be much longer.

@@

More generally, systematically missing variables meant that a trade-off between inclusion of cohort data and appropriate control for confounding was required, resulting in the potential for uncontrolled confounding due to education and BMI. However, sensitivity analysis comparing the commom-set-adjusted (Model 1) and fully-adjusted (Model 2) models in cohorts with a full complement of covariates indicated that further adjustment for BMI and education had minimal impact on the effect estimates.

@@

For the single cohort (Whitehall II), a previously published analysis was available. Tynkkynen _et al._[@tynkkynen2018] analysed the association of blood lipids and risk of all-cause dementia and Alzheimer's disease across several cohorts, including the Whitehall II cohort.

In an attempt to validate my approach, I compared the results from the maximally-adjusted model (Model 2) in this analysis with those reported by Tynkkynen _et al._ for both all-cause dementia and Alzheimer's disease.

The results for the Alzheimer's disease outcome were comparable (Appendix \@ref(appendix-ipd-whitehall)). However, for the vascular dementia outcome (Figure \@ref(fig:whitehallComparisonDementia)), I identified a discrepancy in the association of triglycerides with all-cause dementia estimated by the two analyses. Tynkkynen _et al._[@tynkkynen2018] found a protective effect of triglycerides on this outcome (`r estimate(0.69,0.56,0.85,"HR")`) while this analysis found evidence for a harmful effect (`r estimate(1.26,1.12,1.41,"OR")`).

<!----------------------------------------------------------------------------->
(ref:whitehallComparisonDementia-cap) Comparison of results across the two analysis of the Whitehall II cohort for the all-cause dementia outcome.

(ref:whitehallComparisonDementia-scap) Comparison of results across the two analysis of the Whitehall II cohort

@@

@@

Investigation revealed several possible reasons for this discrepancy. The first is that it was likely to be due to the Whitehall snapshot that each analysis had access to, as comparison of summary statistics indicated that this analysis had substantially more dementia events (N =287 in this analysis vs. n=114 in Tynkkynen _et al._). Additionally, this analysis uses multiple imputation to address missing data, while the Tynkkynen _et al._ does not mention how missing data was addressed, though the reduced precision in the estimates suggest a complete case analysis. Finally, the covariates adjusted for in each analysis were similar, though the previous analysis had access to genetic data, allowing it to adjust for _Apo_$\mathcal{E}4$ status. However, given that _Apo_$\mathcal{E}4$ is a risk factor for increased LDL-c rather than triglycerides,[@bennet2007] it seems unlikely that additional adjustment for this variable is responsible for the discrepancy in findings observed.

@@

While this analysis did not manage to systematically obtain and analyse a large proportion of identified data, a central strength of this analysis is the use of a systematic approach to identify and attempt to contact relevant cohorts. Further, it also enabled incorporation of two previously unanalysed datasets - the CaPS and EPIC Norfolk cohorts - thus providing additional evidence that is used in the triangulation exercise reported in Chapter \@ref(tri-heading).

>>>>

Individual participant data (IPD) meta-analyses are a special type of meta-analysis
in which the authors of relevant studies are contacted with requests to access the raw data on each participant in the original study. This approach is different to the meta-analyses presented in the previous chapter, which used published summary results.

@@

In order to examine the relationship between lipid levels and dementia risk, I applied for access to `r n_applied$total` unique data sources containing information on these variables. However, only a small proportion of these data sources (`r fmt_applied(n_accessed$total)`) provided the requested access.

@@

The reasons for the low response rate to requests for data are explored in this chapter. Finally, I suggest future studies to investigate how data access rates could be improved.

@@

Individual participant data (IPD) meta-analyses are considered to be the gold standard form of evidence synthesis, allowing for the application of a common selection model and analytical approach across all identified cohorts.[@riley2010] They are particularly useful when investigating the impact of participant-level characteristics, something that is not possible with aggregate data unless the results are stratified by the characteristic of interest.[@riley2010;@thompson2005] Knowledge of which groups a treatment will benefit most (or harm least) is a core aim of the move towards personalised medicine.[@riley2020;@hingorani2013]
IPD analyses also offer a mechanism by which previously unanalysed datasets can be incorporated into an analysis, thus expanding the evidence base for a particular research question.

@@

As such, the aims of this analysis are two fold: firstly, to perform an IPD meta-analysis across identified cohorts to examine the impact of participant age-at-measurement and sex on the relationship between lipids and dementia risk; and secondly, to expand the evidence base for the effect of lipids on dementia outcomes by obtaining estimates from previously unanalysed cohorts available via the Dementia Platform UK, a large consortium of dementia cohorts.

@@

@@

Eligible data sources for this analysis were prospective cohort studies (see Section \@ref(ipd-apply-access) for details on how these studies were identified). Data sources which were cross-sectional, either by design or due to the available data (e.g. a study recorded data on participants at multiple time-points or "waves", but only data from a single wave could be accessed) were excluded. Similarly, due to the time and cost restraints within the scope of my thesis project, studies making use of population-level electronic health records were ineligible. These studies are problematic in that they often require extensive project proposals in order to gain access to the data.

No restrictions were put on the number of participants or the length of follow-up, though it was a requirement that participants were dementia free (or assumed to be dementia free, based on age at entry) at baseline.

@@

I considered four blood lipid fractions as part of this analysis, namely: total cholesterol (TC), low-density lipoprotein cholesterol (LDL-c), high-density lipoprotein cholesterol (HDL-c) and triglycerides (TG). Cohorts were eligible for inclusion if they contained data on at least one of these recorded as a continuour variable (i.e. studies with binary "hypercholesteromemia" would be excluded).

@@

### Applying for data access {#ipd-apply-access}

Potentially eligible data sources as defined above were identified via two mechanisms, each with a distinct focus. For both approaches, the number of cohorts responding to the request for data access and, where applicable, the reasons given for a refusal were recorded.

@@

The second approach focused on incorporating relevant, previously-unanalysed data into the analysis, thus providing additional evidence on the relationship between blood lipids and dementia risk. This was achieved through the Dementia Platform UK (DPUK), a collaborative grouping of existing dementia cohorts established by the MRC which works with data owners to make their data readily accessible for secondary analysis.[@bauermeister2020] It provides access to 42 cohorts with over 3 million participants, and makes use of a central streamlined application process for all cohorts with the intent of making it easier to access data from existing data sources.

@@

Where data on one of the exposures (TC, LDL-c, HDL-c, TG) was missing, it was inferred from the other three fractions (where available) using the Friedwald formula (Equation \@ref(eq:total-cholesterol-formula)). Lipids levels were retained as continuous variables rather than dichotomising into a binary hypercholesterolemia exposure, given the additional statistical power this adds to the meta-analysis.[@riley2020;@ensor2018] Where lipid measurements were reported in _mg/dL_, these were converted to _mmol/L_.

@@

Missing data in this analysis was classified as either relating to missing values (a data on a variable of interest was available in a cohort, but some values were missing) or missing variables (a cohort did not collect data on a variable of interest). Variables with missing values were identified, and 20 imputed datasets were created.[@sterne2009] Imputation was performed using the MICE (Multiple Imputation by Chained Equations) using the `mice` package in R.

Missing variables were originally intended to be addressed using a previously published method.[@fibrinogenstudiescollaboration2009] Here, the correlation between the fully-adjusted and partially-adjusted estimate in cohorts with the full set of covariates is used to estimate the fully-adjusted effect in those cohorts missing covariates. However, this method requires several large cohorts to contain the full set of covariates, a condition this analysis failed to meet given the very low response rate. As such, two other common approaches were employed.[@fibrinogenstudiescollaboration2009] In the primary analysis, all cohorts were adjusted for the set of common covariates across cohorts (Model 1: age, sex, smoking, alcohol, diabetes). As a sensitivity analysis, cohorts with the full complement of covariates (Whitehall II and EPIC) were analysed using a maximally-adjusted model (Model 2: Model 1, further adjusted for education, ethnicity, prevalent ischemic heart disease (IHD) and BMI). Results between the common-set-adjusted (Model 1) and maximally-adjusted (Model 2) models were then compared.

@@

In terms of the analytic approach taken, a two-stage IPD analysis was used. Under a two-stage approach, estimates for the effect of each lipid fraction on incident dementia were first calculated for each data source. More specifically, a logistic regression model adjusted for relevant covariates as detailed in the above section was used to quantify the effect of a 1 _mmol/L_ increase in each lipid fraction on each dementia outcome.

Results were expressed as odds ratios (OR). Examination of the data available via the DPUK indicated that the common time-to-event approach used in studies of dementia outcomes would be precluded by the absence of detailed time-to-event data. An overall effect estimate was then produced by combining data-source-specific estimates in a random-effects meta-analysis (see Section \@ref(meta-analysis-methods) for a broader discussion of meta-analysis methods).

@@

A two-stage IPD approach was employed for a number of reasons. Firstly, and most importantly, a two-stage approach allows for siloed data, enabling researchers who are unwilling/unable to provide their data to btain the effect estimates themselves, following a specified analysis plan, and share these with the IPD team.[@riley2010] Secondly, it allows for the production of forest plots of within-cohort estimates, something which is useful for the triangulation exercise in Chapter \@ref(tri-heading). Finally, a two-stage approach is simpler to model, as it uses standard well-documented summary effect estimate meta-analysis techniques, and automatically accounts for methodological issues such as clustering within cohorts[@abo-zaid2013] and the potential for ecological bias.[@burke2017]

@@

In order to investigate the interaction of participant-level characteristics (age and sex) with lipid levels, lipid-covariate interaction terms were extracted and synthesised using a fixed effects meta-analysis.[@fisher2017] All interaction analysis were performed using the common-set-adjusted model (Model 1). To avoid ecological bias, where the between-study association does not reflect the within-study associations, cohorts where there was no within-study variation in the covariate of interest were excluded from the interaction analysis for that covariate.[@burke2017] As an example, it is impossible to estimate the impact of sex on the lipid/dementia relationship in a study which contains only female participants.

<!-- TODO Explain why a fixed effect model was used. -->

@@

Of the `r n_applied$total` studies to which I applied for data access, only three (`r paste0(comma((n_accessed$total/n_applied$total)*100),"%")`) were included in the final analysis. Figure \@ref(fig:cohortFlowchart) details whether the cohorts eventually included in the review were identified by the systematic review or via the DPUK portal. In addition, the reasons for cohorts not being included in the analysis are presented, where available, stratified by application approach.

@@

(ref:cohortFlowchart-cap) __Flowchart detailing sources of data for the IPD analysis__ The number of cohorts approached and the reasons for exclusion are summarised, stratified by identification method (systematic review vs DPUK).

(ref:cohortFlowchart-scap) Flowchart detailing sources of data for the IPD analysis

@@

The European Prospective Investigation of Cancer (EPIC) - Norfolk is a population-based cohort, containing men and women recruited from 35 general practices in Norfolk between 1993 and 1998.[@riboli1997; @riboli2002] Data provided by this cohort for the analysis, dementia was ascertained at at the 5th Health Checkup (2016-2018), providing between 18-25 years of follow-up. The added evidential value of the EPIC cohort is small, given the fact that the data obtained contains only 8 dementia events. The cohort contains no information on dementia subtype, while all covariates of interest bar _Apo_$\mathcal{E}4$ were available.

@@

The Whitehall II study is a prospective cohort study of men and women recruited between 1985 and 1989 from the civil service in London.[@marmot2005] Lipids measurements were available from the third wave of the study, conducted between 1991-1993. The cohort is linked with the Hospital Episode Statistics (HES) database, a database containing details of participant events at NHS hospitals in England, which was used to define covariates at baseline and dementia outcomes.[@zotero-15403] HES data for dementia outcomes was available up to March 2015, providing between 22-24 years of follow-up. Information was available on dementia type, which was classified as all-cause dementia, Alzheimer's disease or vascular dementia. This data source contained details on all covariates of interest except for _Apo_$\mathcal{E}4$ status.

Of note, the Whitehall II cohort was analysed in one of the included studies identified by the systematic review presented in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading),[@tynkkynen2018] meaning that a comparison between the published result and the analysis reported here was possible.

@@

A key consideration in the definition of covariates across cohorts was the classification of age. The Whitehall II study, the largest cohort to which I had access, only shared age data in five-year age bands (e.g. 40-44, 45-49, etc.). To ensure comparability across the cohorts, I created identical categories in the CaPS and EPIC data. This grouped age variable was then used in all subsequent analyses.

@@

<!-- TODO Note here about how the IHD variable in CAPS has a note saying there might be an issue with the variables, while the Whitehall study uses narrow range of codes to define CHD. EPIC is self-reported angina/mi/arrythmia -->

@@

The results from the main effect analysis across the varying lipid fractions on each dementia outcome considered can be seen in Figures \@ref(fig:mainEffectDem) & \@ref(fig:mainEffectVad), respectively. There was weak evidence for an association of any lipid level with either all-cause or vascular dementia, with the exception of a harmful association between raised triglycerides and vascular dementia (`r ipd_vasdem`). For the sole cohort containing data on the Alzheimer's disease outcome (Whitehall II), there was weak evidence for an association of this outcome with any lipid fraction (results shown in comparison with previous analysis of this cohort in Figure \@ref(fig:whitehallComparisonAd-cap))

@@

Given the minimal effect of further adjustment in the maximally-adjusted model, the common-set-adjusted model (Model 1) was used to perform the interaction analyses.

The maximally-adjusted model (Model 2) could have been employed in the analysis of the effect of sex, as the CaPS cohort contains only men and had to be excluded from this analysis, leaving the EPIC and Whitehall cohorts, both of which had a full complement of covariates. However, I decided to employ the same underlying model across the interaction analyses to aid comparability between the age and sex interaction estimates.

@@

<!----------------------------------------------------------------------------->
(ref:interactionDementiaAge-cap)  __Meta-analysis of age-lipid interaction terms for all-cause dementia__ - Age was grouped into 5-year age bands, and the effect estimatse presented represent the OR per 1-step increase in age group. Estimates were obtained using the common-set-adjusted model (Model 1: age, sex, smoking, alcohol, diabetes).

(ref:interactionDementiaAge-scap) Meta-analysis of age-lipid interaction terms for all-cause dementia

@@

<!----------------------------------------------------------------------------->
(ref:interactionDementiaSex-cap) __Meta-analysis of sex-lipid interaction terms for all-cause dementia__ Estimates were obtained using the common-set-adjusted model (Model 1: age, sex, smoking, alcohol, diabetes) and refer to the effect of male gender on the lipid/all-cause dementia association.

(ref:interactionDementiaSex-scap) Meta-analysis of sex-lipid interaction terms for all-cause dementia

@@

<!----------------------------------------------------------------------------->
(ref:interactionVascularAge-cap) __Meta-analysis of age-lipid interaction terms for vascular dementia__ - Age was grouped into 5-year age bands, and the effect estimatse presented represent the OR per 1-step increase in age group. Estimates were obtained using the common-set-adjusted model (Model 1: age, sex, smoking, alcohol, diabetes).

(ref:interactionVascularAge-scap) Meta-analysis of age-lipid interaction terms for vascular dementia

@@

This analysis requested data from `r n_applied$total` data sources, but only obtained data from three, all of which were based in the United Kingdom. No evidence for an effect of lipids on the risk of dementia or related outcomes was identified, except for a harmful association of raised triglycerides with risk of vascular dementia. Similarly, there was weak evidence for an interaction of the effect of lipid levels on dementia outcomes with participants' age, grouped into 5 year bands, or sex.

@@

The obvious key limitation of this analysis is the very low response rate to requests for data access, which may bias the results if there are systematic differences in the association of lipids with dementia outcomes between cohorts that share data and those that do not.[@ahmed2012] Whether or not to press ahead with an IPD analysis in the absence of all (or even most) data is a personal decision, and some previous analyses have highlighted where they decided not to pursue an IPD analysis.[@jaspers2014] For the purposes of this thesis, the decision was made to conduct the IPD analysis, as it provided training in application of the methods in addition to providing new evidence that will be incorporated into the triangulation exercise detailed in Chapter \@ref(tri-heading).

@@

A low response rate is not unexpected, given that a review of IPD studies published between 1987 and 2015 found that fewer than half managed to obtain data from greater than 80% of studies, and that in many cases, the exact percentage of studies for which data was obtained was not accurately reported.[@nevitt2017] However, it is assumed that the ~10% response rate encountered in this analysis is at the lower end of the scale.

There are many likely reasons for this low response rate to requests for data access. In general terms, there are several well-documented barriers that prevent data from being made readily available, including concerns regarding participant privacy, fear of "scooping" or "parasitic" behaviour, and a lack of trust between primary and secondary researchers.[@vanpanhuis2014] More specific to this analysis, IPD meta-analysis including studies other than randomised controlled trials have less success in obtaining IPD from previous published studies.[@nevitt2017] Additionally, while no evidence is available on whether the characteristics of the researcher who is requesting data access influences the response rate and eventual decision, there is the possibility that my position as a PhD student meant I was less likely to elicit a positive response than a well-known senior academic might. The timing of the requests for data access, coinciding with a global pandemic, may also have affected the response rate as researchers prioritised COVID-related work.

Finally, in investigating the potential reasons for the low response rate, I discovered that the method of contact used (email) has been shown to be less successful in eliciting responses from authors when compared with telephoning (which was not used in this analysis).[@danko2019] One of the reasons for this may be that the email addresses reported on publications are more likely to be out of date for older publications. Anecdotally, post-hoc investigation of a subset of cohorts revealed that several corresponding/first authors were no longer at the same institution as when the study was reported, and as a result, were unlikely to have access to the institutional email address listed on the study publication. Despite attempts to track authors as they move between institutions, out-of-date contact details may have contributed to the low response rate.

The obstacles to data access described above are in theory what the DPUK was built to address. However, even with the help of the streamlined application process afforded by the DPUK, accessing sufficient data was a challenge. The response rate among DPUK cohorts a year after application was just 50%. In addition, some cohorts responded saying that the proposed study question was already under investigation by another group, and that they would not share the data on this basis. In light of this, a centralised database of ongoing analyses being performed using DPUK data would be of enormous help, reducing research waste and providing opportunities for collaboration. Finally, the DPUK process would be aided by a clearer distinction between those cohorts that are "DPUK native" (i.e. where a copy of the data is already held on DPUK servers) versus externally hosted, as the response time for externally hosted cohorts is likely to be much longer.

@@

More generally, systematically missing variables required a trade-off between inclusion of cohort data and appropriate control for confounding. The final choice of the common-set-adjusted model for use in the interaction analyses mean there is the potential for residual confounding. However, sensitivity analysis comparing the common-set-adjusted (Model 1) and fully-adjusted (Model 2) models in cohorts with a full complement of covariates indicated that further adjustment for BMI, ethnicity, education and prevalent IHD had minimal impact on the effect estimates. Of note, this is comparable with the analysis of the CPRD data presented in the previous chapter, where adjustment for variable beyond age and sex had a limited impact on the observed effect estimates (see Section \@ref(cprd-impact-additional-covar)).

@@

For the single cohort (Whitehall II), a previously published analysis was available. Tynkkynen _et al._[@tynkkynen2018] analysed the association of blood lipids and risk of all-cause dementia and Alzheimer's disease across several cohorts, including the Whitehall II cohort. In an attempt to validate my approach, I compared the results from the maximally-adjusted model (Model 2) in this analysis with those reported by Tynkkynen _et al._ for both all-cause dementia and Alzheimer's disease.

The results for the Alzheimer's disease outcome were comparable (Figures \@ref(fig:whitehallComparisonAd))). However, for the vascular dementia outcome (Figure \@ref(fig:whitehallComparisonDementia)), I identified a discrepancy in the association of triglycerides with all-cause dementia estimated by the two analyses. Tynkkynen _et al._[@tynkkynen2018] found a protective effect of triglycerides on this outcome (`r estimate(0.69,0.56,0.85,"HR")`) while this analysis found evidence for a harmful effect (`r estimate(1.26,1.12,1.41,"OR")`).

<!----------------------------------------------------------------------------->
(ref:whitehallComparisonDementia-cap) Comparison of results across the two analyses of the Whitehall II cohort for the all-cause dementia outcome.

(ref:whitehallComparisonDementia-scap) Comparison of results across the two analyses of the Whitehall II cohort

@@

<!----------------------------------------------------------------------------->
(ref:whitehallComparisonAd-cap) Comparison of results across the two analyses of the Whitehall II cohort for the Alzheimer's disease outcome.

(ref:whitehallComparisonAd-scap) Comparison of results across the two analyses of the Whitehall II cohort

```{r whitehallComparisonAd, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:whitehallComparisonAd-cap)', out.width='100%', fig.scap='(ref:whitehallComparisonAd-scap)'}
knitr::include_graphics(file.path("figures/ipd/whitehall_comparison_ad.png"))
```
<!----------------------------------------------------------------------------->

@@

Investigation revealed several possible reasons for this discrepancy. The first is that while this analysis uses multiple imputation to address missing data, the Tynkkynen _et al._ analysis does not describe how missing data was handled. However, the reduced precision in the reported estimates suggest a complete case analysis. This interpretation is supported by a comparison of summary statistics, which illustrated that my analysis had substantially more dementia events (N =287 in this analysis vs. n=114 in Tynkkynen _et al._), suggesting that incomplete records were discarded in the previous analysis. Finally, the covariates adjusted for in each analysis were similar, though the previous analysis had access to genetic data allowing it to adjust for _Apo_$\mathcal{E}4$ status. However, given that _Apo_$\mathcal{E}4$ is a risk factor for increased LDL-c rather than triglycerides,[@bennet2007] it seems unlikely that additional adjustment for this variable is responsible for the discrepancy in findings observed.

@@

While this analysis did not manage to obtain and analyse a large proportion of identified data, a central strength of this analysis is the use of a systematic approach to identify and attempt to contact relevant cohorts. Further, it also enabled incorporation of two previously unanalysed datasets - the CaPS and EPIC Norfolk cohorts - thus providing additional evidence that is used in the triangulation exercise reported in Chapter \@ref(tri-heading).

A further strength is the ability of this analysis to investigate the effect of participant characteristics (namely age and sex) on the association of blood lipids and dementia outcomes. To the best of my knowledge, no previous review has examined the interaction of these factors with the observed associations.
